The dynamics of risk perception in a Mediterranean agroecosystem \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-02-07 manuscript/manuscript.tex
pull analysis into new manuscript and combine with manuscript text

---
title: The dynamics of risk perception in a Mediterranean agroecosystem
subtitle: 
titlerunning: The dynamics of risk perception in a Mediterranean agroecosystem
authorrunning: Gauthier
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Nicolas Gauthier 1
  address: Laboratory of Tree-Ring Research & School of Geography and Development, University of Arizona
  email: ngauthier@email.arizona.edu

keywords:
- key
- dictionary
- word

#PACS: 
#- PAC1
#- superPAC

MSC:
- MSC code 1
- MSC code 2

abstract: |
  Small-scale agriculturalists in the Mediterranean Basin rely on multiple strategies including diversification, intensification, and storage to maintain a stable food supply in the face of environmental uncertainty. Each of these strategies requires farmers to make specific resource allocation decisions in response to environmental risks and is thus sensitive to variability in both the spatiotemporal pattern of risk and the ability of farmers to perceive that pattern. In this chapter, I present an agent-based model of a Mediterranean agroecosystem. By driving the model with realistic environmental dynamics derived from simulations of mid-Holocene Mediterranean climate, and by allowing the psychology of risk perception to vary among individual farmers, I explore the hidden vulnerabilities of traditional risk-management strategies to periods of rapid climate change. I show that even when farmers are able to manage risk “optimally” in light of past experience, unanticipated changes in the spatiotemporal pattern of rainfall can still lead to major food shortfalls.

bibliography: bibliography.bib
bibstyle: spbasic
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(raster)
library(tidyverse)
```

```{r}
set.seed(1111)
```

# Introduction {#intro}

Over the past 10,000 years, subsistence farmers in the Mediterranean basin have developed sophisticated strategies to maintain stable food supplies given uncertain rainfall. Their suite of strategies include practices like crop diversification, storage, and exchange [@Halstead1989]). These risk-management strategies all represent forms of \emph{regulatory feedback} used to manage complex socio-ecological systems. Regulatory feedbacks convert a system's high-variance inputs into low-variance outputs by dynamically monitoring the outputs and adjusting internal properties of the system [@Anderies2015]. Risk-managing strategies require farmers to make specific resource allocation decisions in response to environmental risks, and as regulatory feedbacks they are thus sensitive to variability in both the spatiotemporal patterns of risk and the ability of farmers to perceive and act on those patterns.

Crop diversification is an excellent example of a widespread and effective regulatory feedback strategy (Figure 1). Relying on a mix of food types with different climatic tolerances is an efficient way to maintain a robust food supply [@helmers2001separating,Anderies2006]. In the Mediterranean, land-use strategies involving a diversified portfolio of wheat and barley have been employed by even the earliest sedentary farmers, and continue to be used to this day [@GOULD1963a,Slafer1999,Abbo2009a,Marston2011190]. Wheat generally has higher yields but is sensitive to water deficits, while barley has lower yields and is drought tolerant. Planting a mix of wheat and barley, either in the same plot or in a combination of plots, is an effective means of diversifying grain supplies. By adjusting the ratio of wheat to barley, farmers can adapt to changing drought risks.

@@

1. What was the temporal pattern of climate and climate change during the Holocene in the eastern Mediterranean?
2. How well could crop diversification strategies cope with this pattern, given farmers’ imperfect risk perception and decision-making?

Questions:
How would the earliest farmers have made decisions (dealt with uncertainty?) about crop diversification?
ie what were the different algorithms? -- better understand the decision context
What are the consequences of theses different decisions (i.e. how well did they perform) at different points in the history of the mediterranean
How were farmers able to detect changes in the environment, rather than unconsciously adapt to them?
what are the different consequences we'd expect for each
THIS is the most important question because with it I can frame the results around and individual's perception of droughts in the neolithic, and the populaiton-level distribution of drought perceptions at given times

I couple a climate model with high temporal resolution to a simple game theoretic model of agricultural decision making under uncertainty, in order to examine the performance of optimal and suboptimal wheat-barley diversification practices. I model the year-to-year crop diversification strategies made by farmers as an iterated game of fictitious play against nature [@GOULD1963a]. Farmers are boundedly rational, in that they seek to maximize their objective functions given the perceived probabilities of different actions by nature, but work with flawed mental models of climate-related risks due to imperfect recall of past events. Furthermore, by using a climate model to represent the actions of nature, I can more precisely capture the characteristic autocorrelated patterns of rainfall variability, rather than simply drawing rainfall values from a static distribution. This framework allows me to address the sensitivity of diversification strategies to changes and climatic variability and imperfect human perceptions of those changes.

# Decision-making in a game against nature {#sec:1}

The basic decision-making problem facing a farmer seeking to diversify their crops can be thought of as a game against nature.
We can represent this simplified decision context as a "game" in a game theorietic context.
cite luce raiffa 1989 for game against nature, and milnor 1952, Agrawal1968, cassidy1971, gould 1963
The "game" in this context is the farmer's decision of which crops to plant and in what proportions, given uncertianty in the future "state-of-nature", and nature varies between several possible states such as dry and wet years.
payoff matrix goes here
in reality continuous, but cite behavioral econ for people often intuitively solve an easier problem when faced with a complex real-world situation
The exact values in the payoff matrix here are less important than the relative payoffs in each quadrant.
We can think of this game as being the culturally-inherited object. our question is how do individuals make decisions here?
so in this case the cultural norm would be to mix wheat and barley, but the learning is in the exact ratio -- both individual and social learning influence perceptions of eindividual vs social learning tucker 2007

Estimates of yield volume (t/ha) for each crop type in wet and dry years were derived from isotopic studies of ancient wheat and barley samples from archaeological sites in the Mediterranean [@Slafer1999] (Table 1).

@@

```{r}
wheat_dry <- 0
wheat_normal <- 1.6

barley_dry <- 0.93
barley_normal <- 1.18
```

@@

```{r}
expected_yield_wheat <- function(risk) {
  risk * wheat_dry + (1 - risk) * wheat_normal
}

expected_yield_barley <- function(risk) {
  risk * barley_dry + (1 - risk) * barley_normal
}

expected_yield <- function(risk) {
  expected_yield_wheat(risk) + expected_yield_barley(risk)
}
```

```{r, echo = FALSE, fig.cap = 'Expected wheat and barley yields under increasing drought risk with the point of indifference highlighted.'}
indiff <- tibble(risk = (barley_normal - wheat_normal) / (wheat_dry - wheat_normal - barley_dry + barley_normal),
                 yield = risk * wheat_dry + (1 - risk) * wheat_normal)
indiff_label <- paste0('(', round(indiff$risk, 2), ', ', round(indiff$yield, 2), ')')

tibble(risk = seq(0, 1, .1),
      Barley = expected_yield_barley(risk),
      Wheat = expected_yield_wheat(risk)) %>%
  gather(type, yield, Barley:Wheat) %>%
  ggplot(aes(risk, yield)) +
  geom_hline(yintercept = indiff$yield, linetype = 2)+
  geom_vline(xintercept = indiff$risk, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = indiff, color = 'black', size = 3) +
  geom_text(data = indiff, label = indiff_label, nudge_x = 0.1, nudge_y = 0.1) +
  theme_classic() +
  scale_color_discrete(name = 'Crop') +
  labs(x = 'Drought probability', y = 'Expected yield')
```

If an individual can learn about their environment, such as the risk of drought in any given year, then they can behave more rationally by trying to maximize their subjective expected utility
If the probability distributin of nature's moves is known, the farmer can choose the crop mix that simply maximizes the expected crop yields given the empirical frequency distirubtion of nature's moves.
In the language of game theory, this strategy is knoan as a "game of ficticious play" against nature.
This is an effective risk-managing strategy, but like all similar strategies that adapt to a specific pattern of variability, it is vulnerable to changes in the pattern of variability (Janssen eta al 2007)
This strategy works well when the environment is stationary -- nature plays from a fixed probability distirubtion, but is vulnerable to environmental nonstationarity.
That is, when the mean or higher order moments of the rainfall distribution shift, playing a game of fictitious play can backfire because you're too stuck in your ways while the world changes around you.
Thus it is important to understand the dynamics of risk perception -- how subjective risks, and thus subjective expected utility -- rise and fall in uncertain environments.

# The dynamics of risk perception {#sec:2}

```{r}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

tibble(year = seq(5, 50, 5),
       risk = .5) %>%
  mutate(belief = map2(risk, year, ~prior(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(group = year, color = year), size = 1.2) +
  scale_color_viridis_c() +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')

tibble(year = 50,
       risk = seq(.1,.9, .1)) %>%
  mutate(belief = map2(risk, year, ~prior(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(group = risk, color = risk), size = 1.2) +
  scale_color_viridis_c() +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')
```

# Drought risks in the middle Holocene

# Climate Risks and Risk Perception

Using the bias-corrected climate model output, I divided each model year into dry years and wet years. A dry year was any year where less than 300mm of rain fell during the wet season (October-March), the threshold below which wheat crops will generally fail [@Wilkinson1997], and a wet/normal year was defined as any year above this threshold.

@@

\begin{equation}
    P_{dry} = \frac{\sum_{n=t-1}^{t-50} precip_n < 300}{50},
\end{equation} 
where $t$ is the current time step. The 50 year time span was selected to approximate the accumulated observational knowledge of an individual farmer and their immediate household. As a result, this approach does not allow for accumulated social learning, although it could easily be extended to do so in future studies.

## Climate Modeling and Bias Correction

Estimates of drought risk in the eastern Mediterranean during the the past 10,000 years were derived from outputs of the TraCE-21k paleoclimate simulation available on the National Center for Atmospheric Research's Earth System Grid repository (https://www.earthsystemgrid.org). TraCE-21K is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model (GCM) to recreate the transient response of the global climate system to orbital parameters and greenhouse gasses over the past 22,000 years from the Last Glacial Maximum to the present (Figure 2) [@He2011]. It generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of climate forcings (e.g. orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six hourly timescale, and model outputs are archived at a monthly resolution.

TraCE-21k simulation outputs for the past 10,000 years were bias-corrected using the CDF-t method and observed precipitation from a weather station in near the town of Salihli in western Turkey. This location was selected due to its long (~70yr) observational record, the representativeness of western Anatolian climate for the greater eastern Mediterranean, and its proximity to archaeological sites with comparative evidence of crop diversification strategies.

@@

```{r}
konya <- matrix(c(32.5, 37.9), nrow = 1)
#konya <- matrix(c(33.5138, 36.2765), nrow = 1)
```

```{r, cache = TRUE}
precc <- map(list.files('../Data/PRECC', full.names = TRUE)[-12],
             ~brick(.) %>%
               raster::extract(konya, df = TRUE) %>%
  select(-ID) %>%
  gather(time, precc) ) %>%
  bind_rows()
precl <- map(list.files('../Data/PRECL', full.names = TRUE)[-12],
             ~brick(.) %>%
               raster::extract(konya, df = TRUE) %>%
  select(-ID) %>%
  gather(time, precl) ) %>%
  bind_rows()

sim_past <- left_join(precc, precl, by = 'time') %>%
  mutate(rainfall = (precc + precl) * 2.628e+9,
         year = rep(-10200:-5001, each = 12), # add hardcode dates instead of using ncdf time because of date shift with averaging period
         month = rep(1:12, times = 5200)) %>%
  select(-time, -precl, -precc) %>%
  mutate(water_year = if_else(month %in% 10:12, year + 1L, year)) %>% 
   filter(!(month %in% c(6:9))) %>%
  group_by(water_year) %>%
  add_tally() %>%
  filter(n == 8) %>%
  summarise(precip = sum(rainfall)) %>%
  mutate(drought = precip < 300)
```

```{r, fig.height =2, fig.width = 6, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, size = 1) +
  geom_col()  +
  scale_fill_distiller(palette = 'YlOrRd', direction = 1) +
  scale_color_distiller(palette = 'YlOrRd', direction = 1)+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

# The Model
```{r}
agents <- tibble(id = 1:150, 
                 alpha = 1.5,
                 beta = 1.5,
                 risk = alpha / (alpha + beta),
                 age = sample.int(80, size = 150, replace = TRUE),
                 crop = 'barley')
```

```{r}
learn1 <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

learn2 <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = if_else(payoff < .5, 1, drought + alpha),
           beta = if_else(payoff < .5, 1, beta + 1 - drought),
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

# or keep risk parameter constant and reduce the effective sample size, in other words keep the ratio of alpha to beta but decrease the absoulte value
```

Run an experiment. Simulate a time series with varying drought frequency. This is a stepwise nonstationary process
```{r}
droughts <- c((runif(25) < .1),(runif(25) < .5), runif(50) < .2)
```

```{r}
prior2 <- function(a, b) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, a, b)
  tibble('x' = domain, 'y' = val)
}

accumulate(droughts, 
           ~mutate(.x, alpha = .y + alpha, beta = beta + 1 - .y), 
           .init = tibble(alpha = 1.5, beta = 1.5)) %>%
  bind_rows(.id = 'time') %>%
    mutate(time = as.numeric(time), belief = map2(alpha, beta, ~prior2(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(color = time, group = time), size = .8) +
    geom_vline(xintercept = c(0.1, 0.5, 0.2), linetype = 2, color = 'red', size = .8) +
  scale_color_viridis_c() +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')
```

```{r}
simulation1 <- accumulate(droughts, ~learn1(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)

simulation2 <- accumulate(droughts, ~learn2(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

@@

```{r}
test1 <- tibble(time = 1:100,
       risk = c(rep(0.1, 25), rep(0.5, 25), rep(0.2, 50)))

ggplot(simulation1, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
ggplot(simulation2, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()

```

```{r}
ggplot(simulation2, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time, crop) %>%
  tally() %>%
ggplot(aes(time, n, fill = crop)) +
  geom_bar(width = 1, stat = 'identity', position = 'fill') +
  scale_fill_manual(values = c('darkgoldenrod', 'lightgoldenrod')) +
  theme_classic()
```

```{r}
simulation2 %>%
  group_by(time, crop) %>%
  tally() %>%
ggplot(aes(time, n, fill = crop)) +
  geom_bar(width = 1, stat = 'identity', position = 'fill') +
  scale_fill_manual(values = c('darkgoldenrod', 'lightgoldenrod')) +
  theme_classic()
```

@@

Note in these plots above and below we're summing all the wheat and barley in the population, compared to below which shows the wheat proportions of individuals
```{r}
ggplot(simulation1, aes(time, wheat_prop)) +
  geom_point(alpha = .3) +
  theme_classic()

ggplot(simulation2, aes(time, wheat_prop)) +
  geom_point(alpha = .3) +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time) %>%
  summarise(famine = sum(payoff == 0)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
simulation2 %>%
  group_by(time) %>%
  summarise(famine = sum(payoff == 0, na.rm = TRUE)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
```

```{r}
simulation1 %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()

simulation2 %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff, na.rm = TRUE)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()
```

```{r}
library(ggridges)
simulation1 %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(wheat_prop, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()

simulation2 %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(wheat_prop, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()
```

make this a function to calculate payoffs for all wheat or all barley strategies as a comparison
```{r}
tibble(drought = droughts,
       barley = if_else(droughts == FALSE, barley_normal, barley_dry),
       wheat = if_else(droughts == FALSE, wheat_normal, wheat_dry),
       equal = (barley + wheat) / 2) %>%
  mutate(time = 1:n()) %>%
  gather(crop, yield, barley:equal) %>%
  ggplot(aes(time, yield, group = crop, color = crop)) +
  geom_point()
```

```{r}
tibble(drought = droughts,
       barley = if_else(droughts == FALSE, barley_normal, barley_dry),
       wheat = if_else(droughts == FALSE, wheat_normal, wheat_dry),
       equal = (barley + wheat) / 2) %>%
  mutate(time = 1:n()) %>%
  gather(crop, yield, barley:equal) %>%
  mutate(famine = yield == 0) %>%
  ggplot(aes(time, famine, group = crop, color = crop)) +
  geom_line()
```

@@

# References

>>>>

% !TeX program = pdfLaTeX
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

%% load any required packages here

\begin{document}

\title{The dynamics of risk perception in a Mediterranean agroecosystem \thanks{Grants or other notes about the article that should go on the front page
should be placed here. General acknowledgments should be placed at the
end of the article.} }

    \titlerunning{Risk perception in a Mediterranean agroecosystem}

\author{  Nicolas Gauthier 1 \and  }

    \authorrunning{ Gauthier }

\institute{
        Nicolas Gauthier 1 \at
     Laboratory of Tree-Ring Research \& School of Geography and Development,
 University of Arizona \\
     \email{\href{mailto:ngauthier@email.arizona.edu}{\nolinkurl{ngauthier@email.arizona.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Small-scale agriculturalists in the Mediterranean Basin rely on multiple
strategies including diversification, intensification, and storage to
maintain a stable food supply in the face of environmental uncertainty.
Each of these strategies requires farmers to make specific resource
allocation decisions in response to environmental risks and is thus
sensitive to variability in both the spatiotemporal pattern of risk and
the ability of farmers to perceive that pattern. In this chapter, I
present an agent-based model of a Mediterranean agroecosystem. By
driving the model with realistic environmental dynamics derived from
simulations of mid-Holocene Mediterranean climate, and by allowing the
psychology of risk perception to vary among individual farmers, I
explore the hidden vulnerabilities of traditional risk-management
strategies to periods of rapid climate change. I show that even when
farmers are able to manage risk ``optimally'' in light of past
experience, unanticipated changes in the spatiotemporal pattern of
rainfall can still lead to major food shortfalls.
\\
\keywords{
        key \and
        dictionary \and
        word \and
    }

    \subclass{
                    MSC code 1 \and
                    MSC code 2 \and
            }

\end{abstract}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Over the past 10,000 years, subsistence farmers in the Mediterranean
basin have developed sophisticated strategies to maintain stable food
supplies given uncertain rainfall. Their suite of strategies include
practices like crop diversification, storage, and exchange (Halstead and
O'Shea 1989)). These risk-management strategies all represent forms of
\emph{regulatory feedback} used to manage complex socio-ecological
systems. Regulatory feedbacks convert a system's high-variance inputs
into low-variance outputs by dynamically monitoring the outputs and
adjusting internal properties of the system ({\textbf{???}}).
Risk-managing strategies require farmers to make specific resource
allocation decisions in response to environmental risks, and as
regulatory feedbacks they are thus sensitive to variability in both the
spatiotemporal patterns of risk and the ability of farmers to perceive
and act on those patterns.

Crop diversification is an excellent example of a widespread and
effective regulatory feedback strategy (Figure 1). Relying on a mix of
food types with different climatic tolerances is an efficient way to
maintain a robust food supply ({\textbf{???}},Anderies2006). In the
Mediterranean, land-use strategies involving a diversified portfolio of
wheat and barley have been employed by even the earliest sedentary
farmers, and continue to be used to this day (Gould
1963,Slafer1999,Abbo2009a,Marston2011190). Wheat generally has higher
yields but is sensitive to water deficits, while barley has lower yields
and is drought tolerant. Planting a mix of wheat and barley, either in
the same plot or in a combination of plots, is an effective means of
diversifying grain supplies. By adjusting the ratio of wheat to barley,
farmers can adapt to changing drought risks.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  What was the temporal pattern of climate and climate change during the
  Holocene in the eastern Mediterranean?
\item
  How well could crop diversification strategies cope with this pattern,
  given farmers' imperfect risk perception and decision-making?
\end{enumerate}

Questions: How would the earliest farmers have made decisions (dealt
with uncertainty?) about crop diversification? ie what were the
different algorithms? -- better understand the decision context What are
the consequences of theses different decisions (i.e.~how well did they
perform) at different points in the history of the mediterranean How
were farmers able to detect changes in the environment, rather than
unconsciously adapt to them? what are the different consequences we'd
expect for each THIS is the most important question because with it I
can frame the results around and individual's perception of droughts in
the neolithic, and the populaiton-level distribution of drought
perceptions at given times

I couple a climate model with high temporal resolution to a simple game
theoretic model of agricultural decision making under uncertainty, in
order to examine the performance of optimal and suboptimal wheat-barley
diversification practices. I model the year-to-year crop diversification
strategies made by farmers as an iterated game of fictitious play
against nature (Gould 1963). Farmers are boundedly rational, in that
they seek to maximize their objective functions given the perceived
probabilities of different actions by nature, but work with flawed
mental models of climate-related risks due to imperfect recall of past
events. Furthermore, by using a climate model to represent the actions
of nature, I can more precisely capture the characteristic
autocorrelated patterns of rainfall variability, rather than simply
drawing rainfall values from a static distribution. This framework
allows me to address the sensitivity of diversification strategies to
changes and climatic variability and imperfect human perceptions of
those changes.

\hypertarget{sec:1}{%
\section{Decision-making in a game against nature}\label{sec:1}}

The basic decision-making problem facing a farmer seeking to diversify
their crops can be thought of as a game against nature. We can represent
this simplified decision context as a ``game'' in a game theorietic
context. cite luce raiffa 1989 for game against nature, and milnor 1952,
Agrawal1968, cassidy1971, gould 1963 The ``game'' in this context is the
farmer's decision of which crops to plant and in what proportions, given
uncertianty in the future ``state-of-nature'', and nature varies between
several possible states such as dry and wet years. payoff matrix goes
here in reality continuous, but cite behavioral econ for people often
intuitively solve an easier problem when faced with a complex real-world
situation The exact values in the payoff matrix here are less important
than the relative payoffs in each quadrant. We can think of this game as
being the culturally-inherited object. our question is how do
individuals make decisions here? so in this case the cultural norm would
be to mix wheat and barley, but the learning is in the exact ratio --
both individual and social learning influence perceptions of eindividual
vs social learning tucker 2007

Estimates of yield volume (t/ha) for each crop type in wet and dry years
were derived from isotopic studies of ancient wheat and barley samples
from archaeological sites in the Mediterranean (Araus, Slafer, and
Romagosa 1999) (Table 1).

@@

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Expected wheat and barley yields under increasing drought risk
with the point of indifference highlighted.}
\end{figure}

If an individual can learn about their environment, such as the risk of
drought in any given year, then they can behave more rationally by
trying to maximize their subjective expected utility If the probability
distributin of nature's moves is known, the farmer can choose the crop
mix that simply maximizes the expected crop yields given the empirical
frequency distirubtion of nature's moves. In the language of game
theory, this strategy is knoan as a ``game of ficticious play'' against
nature. This is an effective risk-managing strategy, but like all
similar strategies that adapt to a specific pattern of variability, it
is vulnerable to changes in the pattern of variability (Janssen eta al
2007) This strategy works well when the environment is stationary --
nature plays from a fixed probability distirubtion, but is vulnerable to
environmental nonstationarity. That is, when the mean or higher order
moments of the rainfall distribution shift, playing a game of fictitious
play can backfire because you're too stuck in your ways while the world
changes around you. Thus it is important to understand the dynamics of
risk perception -- how subjective risks, and thus subjective expected
utility -- rise and fall in uncertain environments.

\hypertarget{sec:2}{%
\section{The dynamics of risk perception}\label{sec:2}}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-2.pdf}

\hypertarget{drought-risks-in-the-middle-holocene}{%
\section{Drought risks in the middle
Holocene}\label{drought-risks-in-the-middle-holocene}}

\hypertarget{climate-risks-and-risk-perception}{%
\section{Climate Risks and Risk
Perception}\label{climate-risks-and-risk-perception}}

Using the bias-corrected climate model output, I divided each model year
into dry years and wet years. A dry year was any year where less than
300mm of rain fell during the wet season (October-March), the threshold
below which wheat crops will generally fail ({\textbf{???}}), and a
wet/normal year was defined as any year above this threshold.

@@

\begin{equation}
    P_{dry} = \frac{\sum_{n=t-1}^{t-50} precip_n < 300}{50},
\end{equation} where \(t\) is the current time step. The 50 year time
span was selected to approximate the accumulated observational knowledge
of an individual farmer and their immediate household. As a result, this
approach does not allow for accumulated social learning, although it
could easily be extended to do so in future studies.

\hypertarget{climate-modeling-and-bias-correction}{%
\subsection{Climate Modeling and Bias
Correction}\label{climate-modeling-and-bias-correction}}

Estimates of drought risk in the eastern Mediterranean during the the
past 10,000 years were derived from outputs of the TraCE-21k
paleoclimate simulation available on the National Center for Atmospheric
Research's Earth System Grid repository
(\url{https://www.earthsystemgrid.org}). TraCE-21K is a state-of-the-art
simulation that uses a coupled atmosphere-ocean general circulation
model (GCM) to recreate the transient response of the global climate
system to orbital parameters and greenhouse gasses over the past 22,000
years from the Last Glacial Maximum to the present (Figure 2) (He 2011).
It generates physically consistent spatiotemporal climate dynamics,
driven by current best estimates of climate forcings (e.g.~orbit,
greenhouse gasses, glacial meltwater flux). The model simulates these
dynamics on a six hourly timescale, and model outputs are archived at a
monthly resolution.

TraCE-21k simulation outputs for the past 10,000 years were
bias-corrected using the CDF-t method and observed precipitation from a
weather station in near the town of Salihli in western Turkey. This
location was selected due to its long (\textasciitilde{}70yr)
observational record, the representativeness of western Anatolian
climate for the greater eastern Mediterranean, and its proximity to
archaeological sites with comparative evidence of crop diversification
strategies.

@@

\begin{verbatim}
## Loading required namespace: ncdf4
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Annual risk of wheat crop failure due to drought, averaged by
fifty-year period}
\end{figure}

\hypertarget{the-model}{%
\section{The Model}\label{the-model}}

Run an experiment. Simulate a time series with varying drought
frequency. This is a stepwise nonstationary process

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-12-1.pdf}

@@

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-14-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-15-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-16-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-18-1.pdf}

@@

Note in these plots above and below we're summing all the wheat and
barley in the population, compared to below which shows the wheat
proportions of individuals
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-19-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-19-2.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-20-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-20-2.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-21-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-21-2.pdf}

\begin{verbatim}
## Picking joint bandwidth of 0.0146
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-22-1.pdf}

\begin{verbatim}
## Picking joint bandwidth of 0.0204
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-22-2.pdf}

make this a function to calculate payoffs for all wheat or all barley
strategies as a comparison
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-24-1.pdf}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Slafer1999}{}%
Araus, J. L., G. A. Slafer, and I. Romagosa. 1999. ``Durum wheat and
barley yields in antiquity estimated from 13C discrimination of
archaeological grains: a case study from the Western Mediterranean
Basin.'' \emph{Australian Journal of Plant Physiology} 26 (4): 345.
\url{https://doi.org/10.1071/PP98141}.

\leavevmode\hypertarget{ref-Garnsey1989}{}%
Garnsey, Peter. 1989. \emph{Famine and Food Supply in the Graeco-Roman
World: Responses to Risk and Crisis}. Cambridge University Press.
